Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit

### 2023-12-20 Manuscript/Response to the Reviews.Rmd

*Summary: The ciliary photoreceptor cells and its downstream neurons of larval annelid must be orchestrated in a specific pattern to promote downward swimming in response to long duration of UV exposure. The authors first conducted neuroanatomical examination of the circuit to identify NOS-expression neurons (INNOS) that are immediately downstream to the ciliary photoreceptor cells. The INNOS is activated by UV and produces NO. The NOS is required for UV avoidance by Platynereis larvae and neural dynamics of the photoreceptor cells and their downstream circuit. Following up the RNA-seq data with in situ hybridization experiments, the authors found that two unconventional guanylate cyclases, NIT-GC1 and NIT-GC2, are expressed and localized in different subcellular domain of the photoreceptor cells. Experiments using the culture cells and genetically encoded sensors demonstrated that NIT-GC1 can generate cGMP in response to nitric oxide. Finally, authors build a mathematical model that fit the live imaging data and used it to predict how the magnitude of the photoreceptor activation varied by intensity and duration of UV light.*

*Strengths: The authors conducted comprehensive interrogations of the UV avoidance pathway at the molecular and circuit levels, and constructed a mathematical model. The main conclusions are supported by layers of evidence from different assays.*

*Weaknesses: Statistics are missing in both figure legends and methods. The perturbations of genes and molecules were not cell-type-specific and therefore the observed behavioral defect could be attributed to the malfunction of the circuit elsewhere not examined in this study. I suggest adding more explanation about the functions of other NOS-expressing cells and conducting a control experiment to test behavioral response to a non-visual stimulus.*

1.  As you point out, we have performed statistical tests and added explanatory text to figure legends.
2.  Only one NOS homologue has been found in the Platynereis genome. Note that it is specifically expressed in NOS interneurons, which are our target in this study, based on the results of single-cell transcriptome analysis and in situ hybridisation against the NOS gene. However, as you have pointed out, the functions of nitric oxide are diverse and we have revised some of our discussion to include the possibility of its interneurons outside the non-visual system.

@@

The size of the chamber is limited due to the experimental system, so many larvae tend to accumulate at the bottom of the chamber by descending swimming at the end of the experiment. This may have led to variability in the data at the end point and lengthened the error bars.

A 120 s exposure time was most suitable for quantifying larval behaviour from a previous paper by Verasztó et al. 2017.

*Page 13: "During the UV response, prototroch cilia beat slower than trunk cilia, resulting in a head-down stable state ('rear-wheel drive'). In contrast, during the pressure response prototroch cilia beat faster than trunk cilia, leading to a head-up orientation ('front-wheel drive'). Testing this hypothesis will require biophysical experiments and mathematical modelling." Authors should carry out ciliary beating analysis under UV light in the current study with NOS mutant larvae. Since the pressure and UV detection systems are closely related, comparing the difference in ciliary beating is important to demonstrate this hypothesis. Further, did the authors check the Ca sensor GCaMP6s under pressure conditions?*

We have newly analysed the ciliary beat frequency of larvae by UV stimulation and added Figure 3---figure supplement 3.

@@

Recommendations for the authors: please note that you control which revisions to undertake from the public reviews and recommendations for the authors

**Reviewer #1 (Recommendations For The Authors):**

1)  *The current introduction focuses on the nitric oxide signaling. It would be helpful for readers to have an introductory section about Platynereis larvae (a total number of neurons, etc) and rationales to use its nervous system as a model to study mechanisms of NO signaling and gating of visual response.*

    The 3dpf larvae reveals 90 non-neuronal cell types and 180 neuronal cell types, mapping the complete neuronal connectome. This can shed light on the question of how synaptic and non-synaptic networks regulate their various patterns of behaviour.

@@

2)  *The readers would want to know why these larvae swim downward in response to UV and upward to 480nm light. Is that for maintaining the certain depth from the surface of the water?*

    It has been suggested that the ciliary photoreceptor circuit, which senses UV light, and the rhabdomeric photoreceptor circuit, which senses blue light, exchange messages with each other and that the two work together to form a depth gauge. By allowing larvae to swim at their preferred depth, the depth gauge influences where they end up when they become adults.

@@

*Really, that's the only, relatively minor issue I could find. This was a pleasure to read, and I learned a lot. Congratulations on an excellent study.*

**Reviewer #3 (Recommendations For The Authors):**

*Page 3, Figure 1 A: The authors reconstructed the cPRC circuit in 3-day-old larvae and detected NOS gene expression in 2-day-old larvae in Figure 1D&E. Can the authors provide a cPRC circuit reconstruction for 2-day-old larvae?*

We have added further data on NOS gene expression in three-day-old larvae to Figure 1---supplement 2C, although it is difficult for us to show data on the connectome of two-day-old larvae.

@@

*Page 4: Two Platynereis NOS knockout lines (NOSΔ11/Δ11 and Δ23/Δ23) using the CRISPR/Cas9: Could you direct me to the knockout conformation results for NOS knockout lines NOSΔ11/Δ11 and Δ23/Δ23?*

We have shown in Figure 3---supplement 1A.

*Page 4: "Three day-old but not two-day-old NOS-mutant larvae also showed reduced phototactic behaviour, suggesting a function for NOS in the visual eyes that mediate three-day-old phototaxis" This sentence is unclear. Why is it only three days old but not two days old?*

The visual eyes (adult eyes) co-express r-opsin1 and Go-opsin1, which are known to be involved in phototaxis, and it is thought that differences in phototaxis behaviour may arise because the visual eyes themselves are still underdeveloped in two-day-old larvae.

@@

*Page 5: "All trajectories start at 0 x and y position and time 0 corresponding to 10 sec after the onset of 395 nm stimulation from the side." How did the authors determine the 10-second duration? Are there any potential reasons for this choice?*

Due to constraints imposed by the experimental set-up and analysis, we found it difficult to track the same larvae for extended periods of time, so we restricted our analysis to a total of 40 s of tracking, starting 10 s before stimulus onset and continuing for 30 s after stimulus onset.

*"(A) Swimming trajectories of wild type (WT, n=32) and NOS mutant (NOSΔ11/Δ11, n=26 and NOSΔ23/Δ23, n=47) three-day-old larvae." Authors keep shifting between 2-day and 3-day-old larval data in Figures 1, 2, and 3, causing inconsistency.*

Due to sample immobilisation constraints in live imaging, we found it difficult to perform DAF-FM and calcium imaging with three-day-old larvae. However, we have observed similar patterns of UV-avoidance behaviour and NOS gene expression in two-day-old larvae as in three-day-old larvae, and we believe that it is possible to make a correspondence.

@@

*Page 5: "Figure 3. NOS is required for UV avoidance in Platynereis larvae." Why didn't the authors present any statistics on the plots? A statistical test is required to prove that the difference is significant.*

We have added statistical tests to the data in Figure3.

*Page 5: "Analysis of sGCs in Platynereis indicated that these genes are not expressed in any of the cells of the cPRC circuit (not shown and (Verasztó et al., 2017))." Hence the data is relevant; please provide this data in supplementary.*

We re-analysed previous single-cell data (Achim et al., 2018) and found no sGC homologues detected in cPRC and INNOS. However, a few were detected in INRGW. Additions were made to the supplement data regarding these data and the text was also amended.
